---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
1. Exploratory Data Analysis (EDA)


```{r}
# Load necessary libraries
library(ggcorrplot)

library(tidyverse)
```
```{r}
# Read the dataset
data <- read_csv("CityHousing.csv")
```
#display dataset
```{r}
data
```
```{r}
# Display structure and summary
str(data)
summary(data)
```
```{r}
# Scatter Plot: HousePrice vs SquareFootage
plot(data$SquareFootage,data$HousePrice, 
     main = "Scatter Plot: HousePrice vs SquareFootage", 
     xlab = "Square Footage (sq. ft.)", 
     ylab = "House Price (USD)", 
     col = "blue", 
     pch = 19)

```
```{r}
#Histogram: Distribution of HousePrice
hist(data$HousePrice, 
     main = "Histogram of House Prices", 
     xlab = "House Price (USD)", 
     col = "lightblue", 
     border = "black", 
     breaks = 20)

```
```{r}
# Boxplot: HousePrice by Bedrooms
```


```{r}
boxplot(data$HousePrice ~ data$Bedrooms, 
        main = "Boxplot of House Prices by Bedrooms", 
        xlab = "Number of Bedrooms", 
        ylab = "House Price (USD)", 
        col = "lightgreen")

```

```{r}
# Line Plot: DistanceToCityCenter vs HousePrice
plot(data$DistanceToCityCenter, data$HousePrice, 
     type = "l", 
     main = "Line Plot: Distance to City Center vs House Price", 
     xlab = "Distance to City Center (miles)", 
     ylab = "House Price (USD)", 
     col = "green", 
     lwd = 2)

```

```{r}
#identify the missing values
colSums(is.na(data))  # Count missing values in each column

```
no any missing values in this data set




```{r}
# Check the type of square_footage column
str(data$SquareFootage)

# If needed, convert to numeric
data$SquareFootage <- as.numeric(data$SquareFootage)

```

```{r}
# Replace outliers using IQR
Q1 <- quantile(data$squarefootage, 0.25, na.rm = TRUE)
Q3 <- quantile(data$squarefootage, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR





```
```{r}
#Compute correlations
correlation_matrix <- cor(data %>% select(where(is.numeric)), use = "complete.obs")
print(correlation_matrix)

```

```{r}
#Visualize as a heatmap

ggcorrplot(correlation_matrix, lab = TRUE, title = "Correlation Heatmap")

```
          *larger homes tend to cost more
          *more bedrooms might increase price but not always proportionally
          *closer to the city, higher the price
          
          
          
2. Build the Regression Model

```{r}
#Build the regression model
full_model <- lm(HousePrice ~ SquareFootage + Bedrooms + DistanceToCityCenter + NeighborhoodIncome, data = data)

# Summary of the model
summary(full_model)

```
```{r}
#Check for Multicollinearity
library(car)  # For VIF calculation
vif(full_model)

```

```{r}
#Refine the Model
refined_model <- lm(HousePrice ~ SquareFootage + Bedrooms + DistanceToCityCenter + NeighborhoodIncome, data = data)
summary(refined_model)

```
```{r}
#Predict House Prices
predictions <- predict(refined_model, newdata = data)
head(predictions)

```
```{r}
#Evaluate Model Performance
actual <- data$house_price
mse <- mean((actual - predictions)^2)
r_squared <- summary(refined_model)$r.squared

print(paste("MSE:", mse))
print(paste("R-squared:", r_squared))

```

 3. Interpret the Results
 
```{r}
summary(refined_model)

```


(Intercept)
Value: 49,110
When all predictors (SquareFootage, Bedrooms, DistanceToCityCenter, NeighborhoodIncome) are zero, the baseline house price is $49,110.
Note: This is mostly a theoretical reference and not directly meaningful since a house with zero square footage, bedrooms, and income isn’t practical.

SquareFootage (199.5)
Value: 199.5
For every additional square foot of house area, the house price increases by approximately $199.50, keeping all other factors constant.
This strong, positive relationship is logical: larger homes typically cost more.
          
Bedrooms (10,940)
Value: 10,940
Each additional bedroom increases the house price by $10,940, assuming other factors remain constant.
This indicates a significant positive impact of more bedrooms, which adds value to the property.
          
          
DistanceToCityCenter (-1,201)
Value: -1,201
For every additional mile away from the city center, the house price decreases by $1,201, holding all other variables constant.
This negative relationship is expected, as proximity to the city center generally increases property value due to accessibility and convenience.             


NeighborhoodIncome (0.2838)
Value: 0.2838
For every $1 increase in average neighborhood income, the house price increases by approximately 28 cents, holding all other factors constant.
Though the effect is relatively small, it reflects the influence of a wealthier neighborhood on property values.


```{r}
# Install and load the car package
library(car)
```

```{r}
# Calculate the VIF for each predictor in the model
vif(refined_model)

```



4. Validate Model Assumptions 

01.Linearity
```{r}
# Residuals vs Fitted Values
plot(refined_model$fitted.values, residuals(refined_model),
     main = "Residuals vs Fitted Values",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch = 20, col = "blue")
abline(h = 0, lty = 2, col = "red") 

```
*The residuals should scatter randomly around the horizontal line. Any clear patterns (e.g., curves) indicate a violation of linearity


02. Normality of Residuals

```{r}
# Q-Q Plot
qqnorm(residuals(refined_model), main = "Normal Q-Q Plot")
qqline(residuals(refined_model), col = "red")  

```
*Residuals should fall along the diagonal line. Deviations, especially in the tails, suggest a violation of normality.


03. Homoscedasticity

```{r}
#Scale-Location Plot
plot(refined_model, which = 3) 

```
*The residuals should be evenly spread around the horizontal line. A funnel shape (narrowing or widening) indicates heteroscedasticity.

04. Independence of Errors
```{r}
# Durbin-Watson Test
library(lmtest)
dwtest(refined_model)

```

5. Generate Insights and Recommendations

*Insights*

The regression model highlights that square footage and number of bedrooms are the strongest positive predictors of house prices. Larger homes and more bedrooms significantly increase value, with an additional square foot adding about $199.50 and each bedroom contributing $10,940. Conversely, distance from the city center negatively impacts prices, with houses losing $1,201 per mile. While neighborhood income positively influences prices, its effect is relatively minor. The model has a high R-squared, indicating it explains a substantial portion of the variability in house prices, and all predictors are statistically significant.


*Recommendations*

Based on the findings, buyers should prioritize properties with larger square footage and more bedrooms, especially those closer to city centers. Sellers can maximize property value by emphasizing these features in listings. Real estate developers should focus on building family-sized homes in urban or high-income areas to meet market demand. Policymakers should invest in better transportation links to suburban areas to boost the value of properties farther from the city.




